\vspace*{\credgap}
{\noindent\LARGE\sc Letter from the Editor}
\vspace{\credgap}

\vspace{\ackgap}\noindent
Dear Reader,

\vspace{\credgap}\noindent
It is with pleasure that I present the 25\textsuperscript{th} edition
of \emph{Aporia}, the undergraduate philosophy journal of St Andrews.
I hope it shall prove a stimulating and enjoyable read.

My particular thanks, firstly, to Joe Bradstreet, my deputy, who has
been a well-informed and clever editor, a diligent administrator, and
a pleasant balance to all my neuroticisms. My thanks also to Christina
Landys Herre, who bravely shouldered much of the administrative
burden.  And I cannot fail to mention Mohit Agarwal, whose technical
prowess and willingness to take on tasks far beyond my capabilities
have been utterly invaluable. I am extremely grateful to the whole
editorial team for their dedication, expertise, and good humour; I
should have been quite at sea with many of the papers were it not for
their excellent work.

I am often teasingly asked about our readership; I think I may honestly
answer that the people who truly gain the most from \emph{Aporia} are
the editors and writers. From my editorship I have been exposed to areas
of philosophy untouched by my courses, learned more about the rights
(and perhaps more frequently wrongs) of academic writing than any
lecturer has taught me, and acquired the skill of working with a
relatively large and diverse team. For all the vexations of the
position, I am pleased to have had the opportunity of dipping my toe
into the murky waters of academic research and publishing.

It is a real triumph that \emph{Aporia} has survived 18 years of
disorganized philosophy students, incomprehensible technological
advancements, interrupted teaching, and unreliable funding. Let us hope
it shall long continue for the next generations of budding philosophers.

\vspace{\credgap}\noindent
Yours faithfully,

\vspace{\ackgap}\noindent
Kirsty Graham

\noindent
Editor-in-Chief, \emph{Aporia}
